Correcting for bias in high throughput sequencing data

### 2011-06-12 seqbias-paper.tex
redo freqs figure, etc

\address{
$^{1}$Deportment of Computer Science and Engineering, University of
Washington, Seattle, WA 98195-2350, USA\\
$^{2}$Fred Hutchinson Cancer Research Center, Seattle, WA 98109\\
$^{3}$Department of Microbiology, University of Washington, Seattle, WA
98195-7242, USA}

@@

\begin{abstract}
Quantification of RNA abundance in RNA-Seq experiments is often conflated with
protocol specific bias as a result of PCR amplification, differing primer
affinities, and random priming with a non-uniform distribution of primer
sequences.  We propose a method to measure and compensate for these influences
using a simple graphical model, trained without the need of existing gene
annotations.
\end{abstract}

@@

In the last few years, RNA-Seq has emerged as a promising alternative to
microarrays in quantifying RNA abundance. But, as microarray technology has
brought with it technical challenges ranging from robust normalization to
addressing cross-hybridization, RNA-Seq presents a new set of problems.

A significant example is the effect of transcript length and quantification.
As noted by Oshlack, et. al., transcript length biases discovery of differentially
expression, since more reads are aligned to longer transcripts, granting
higher statistical significance in tests differential expression
\cite{Oshlack2009}. Recently, some methods have been developed attempting to account for
this effect in gene set enrichment tests \cite{Young2010} \cite{Gao2011}.
Additionally, differing transcript lengths introduce ``edge
effects'', as very short fragments from the end of a transcript will not be
sequenced. NEUMA \cite{Lee2010} and FPKM \cite{Trapnell2010} are two methods
that attempt to take this into account.

@@

To estimate sequencing bias, we train a Bayesian network classifier in which
each node represents a position in the sequence, relative to the read start, and
edges encode dependency between positions. Bayesian networks have been
applied to recognize motifs in nucleotide sequences in the past, in particular
in modeling splice cites \cite{Cai2000} \cite{Chen2005} and transcription factor
binding sites \cite{BenGal2005} \cite{Grau2006} \cite{Pudimat2005}.

@@

All data sets were mapped using Bowtie \cite{Langmead2009} against the latest
genome assemblies available from the UCSC Genome Browser \cite{Rhead2010} during
preparation.

@@

\section{Availability}

The method is implemented in the R/Bioconductor \cite{Gentleman2004} package \texttt{seqbias}.

\bibliographystyle{plain}
\bibliography{seqbias-paper}

>>>>

\address{
$^{1}$Deportment of Computer Science and Engineering, University of
Washington, Seattle, WA 98195-2350, USA\\
$^{2}$Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA\\
$^{3}$Department of Microbiology, University of Washington, Seattle, WA
98195-7242, USA}

@@

\begin{abstract}

\section{Motivation:}

Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
often conflated with protocal-specific bias, influenced by PCR amplification,
and differing primer affinities and mixtures, for example. The result is
potentially inaccurate quantification, affecting de novo gene annotation and
isoform quantification in RNA-Seq, and peak calling in ChiP-Seq.

\section{Results:}

We present a method to measure and correct for these influences using a simple
graphical model. Our model does not rely on existing gene annotations, making it
applicable to any high throughput sequencing data. We evaluate our method on
several datasets, and by multiple criteria, demonstrating that it effectively
decreases bias and increases uniformity.

\section{Availability:}
The method is implemented in the \texttt{seqbias} R/Bioconductor package,
available freely under the LGPL licence from
\href{http://bioconductor.org}{http://bioconductor.org}.

\section{Contact:}
\href{dcjones@cs.washington.edu}{dcjones@cs.washington.edu}

\end{abstract}

@@

In the last few years, RNA-Seq has emerged as a promising alternative to
microarrays in quantifying RNA abundance. But, as microarray technology has
brought with it technical challenges ranging from developing robust
normalization to accounting for cross-hybridization, RNA-Seq presents a new set
of problems.

A significant example is the effect of transcript length on quantification.
As noted by Oshlack, et. al., transcript length biases discovery of differentially
expression, since more reads are aligned to longer transcripts, granting
higher statistical significance in tests differential expression
\cite{Oshlack2009}. Recently, some methods have been developed attempting to account for
this effect in gene set enrichment tests \cite{Young2010} \cite{Gao2011}.
Additionally, differing transcript lengths introduce ``edge
effects'', as very short fragments from the end of a transcript will not be
sequenced. NEUMA \cite{Lee2010} and FPKM \cite{Trapnell2010} are two methods
that attempt to take this into account.

@@

To estimate sequencing bias, we train a Bayesian network classifier in which
each node represents a position in the sequence, relative to the read start, and
edges encode dependency between positions. Bayesian networks have been
applied to recognize motifs in nucleotide sequences in the past, in particular
in modeling splice cites \cite{Cai2000} \cite{Chen2005} and transcription factor
binding sites \cite{Ben-Gal2005} \cite{Grau2006} \cite{Pudimat2005}.

@@

All data sets were mapped using Bowtie \cite{Langmead2009} against the latest
genome assemblies available from the UCSC Genome Browser \cite{Karolchik2008}
during preparation.

@@

\bibliographystyle{natbib}
\bibliography{seqbias-paper}
